---
title: 'Interactive #SOTEU circle plots and wordclouds'
author: Giorgio Comai
date: '2017-09-22'
slug: interactive-soteu-circle-plots-and-wordclouds
categories:
  - rstats
  - interactive
  - shiny
tags: []
---

In [a previous post](../../13/exploring-soteu-speeches), I described how I created the data visualisations that have been published by [EUobserver](https://euobserver.com/institutional/138997), [VoxEurope](http://www.voxeurop.eu/en/2017/state-european-union-speech-5121369), and [balcanicaucaso.org](https://www.balcanicaucaso.org/eng/Areas/Europe/Counting-the-words-that-count-182535).

Some of the graphs were based on keywords selected by the editors, such as for example this graph:

```{r echo=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman") # for taking care of package installation/loading

pacman::p_load("tidyverse")  # data processing


speeches_by_speech <-  readRDS(file.path("..", "..", "static", "soteu", "speeches_by_speech.rds"))

keywords_circles <- speeches_by_speech %>% 
  filter(word == "solidarity" | word == "refugee" |word == "crisis" | word == "greece"| word == "climate" | word == "citizens"| word == "euro" | word == "jobs" ) %>% 
  mutate(word = factor(word, levels = unique(word))) %>% # makes sure order is kept for graph 
  ggplot(mapping = aes(x = word, y = id, size = n, colour = n, label = n)) +
  geom_point(shape = 21, stroke = 3, size = 10) +
  geom_text(mapping = aes(size = 25), colour = "black") +
  scale_colour_continuous(low = "yellow", high = "red")+
  scale_x_discrete(name = "") +
  scale_y_discrete(name = "") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position="none") +
  labs(title = "Frequency of selected keywords in #SOTEU speeches")
keywords_circles
```

Looking at it, you may have wondered: "but what about this other word?"

## Interactive version of the circle plot

Wonder no more! Now, it's easy to find out through an interactive version of the same graph. You can insert comma-separated words to create your own circle plot based on #SOTEU speeches. As for the original graph, common stopwords (e.g. "and", "or", etc.) and self-referential terms ("european", "commission", etc.) have been excluded. Terms should all be lower-case, and to facilitate meaningful comparisons this interactive defaults to *&ast;term&ast;*, so the results of "example" would correspond to the the sum of occurrences of "example" and "examples". ^[This is mostly what you want, but beware, in particular with short words: "eco" would present the sum of all occurrences of "economic", "ecologic", etc.]

<iframe src="http://cloud.giorgiocomai.eu:3838/SoteuCircles/", width="660" height="660" frameBorder="0"></iframe>

## Thematic wordclouds

The final version of the graphs published included a number of wordclouds, but not a type of wordcloud that I thought was particularly interesting. A wordcloud based on all the words found in the same sentence as a given keyword. For example, what words are found in the same sentence as "crisis"?

```{r echo=FALSE, message=FALSE}
pacman::p_load("tidyverse")  # data processing
pacman::p_load("tidytext")  # text processing
pacman::p_load("stringr") # text processing
pacman::p_load("wordcloud") # exactly

speeches_by_sentence <-  readRDS(file.path("..", "..", "static", "soteu", "speeches_by_sentence.rds"))

exclude <- data_frame(word = c("european", "europe", "union", "commission", "eu", "means", "europe's"))

speeches_by_sentence %>% 
  filter(stringr::str_detect(string = sentence, pattern = "crisis")) %>% 
  unnest_tokens(input = sentence, output = word) %>% 
  anti_join(stop_words) %>% 
  anti_join(exclude) %>% 
  count(word) %>%   with(wordcloud(word, n, scale = c(2.5, 1), max.words = 50, 
                 random.order = FALSE, vfont=c("sans serif","plain")))
```

But hey, what if you're interested in some other keywords? There you go. You may want to try *russia*, *balkan*, or whatever you are interested in.

<iframe src="http://cloud.giorgiocomai.eu:3838/SoteuThematicWordclouds/", width="660" height="660" frameBorder="0"></iframe>
