---
title: "Animating 'One Degree Warmer' time series with ggplot2 and gganimate"
author: 'Giorgio Comai'
date: '2018-10-12'
slug: one-degree-warmer-animating-time-series-with-ggplot-and-gganimate
categories: [rstats]
tags: []
---

[__Notice: this is still a draft__]

```{r message=FALSE}
if(!require("pacman")) install.packages("pacman")

knitr::opts_chunk$set(echo = TRUE)

# pacman::p_load("googlesheets") # to get the data from google docs
pacman::p_load("readxl") # to read xlsx file
pacman::p_load("ggrepel") # used for labeling
  
if (!require("gganimate")) devtools::install_github("thomasp85/gganimate")
library("gganimate") # get moving!


pacman::p_load("tidyverse") # all the goodies!

# If you want to change font
# pacman::p_load("extrafont")
# extrafont::loadfonts()
```



# Get and clean the data



```{r message=FALSE}
# indicators
# https://docs.google.com/spreadsheets/d/1X5J3GbAKGL5DxwB0aUzAutslHgZ7XFazkVmfQyXhIqM/edit#gid=595233479

# Temperature per year and location
#https://docs.google.com/spreadsheets/d/1avuUJmVNljyb2uotHKDGgrG8T74O7CcwG15p0budkH8/edit?usp=sharing

dir.create(path = 
             file.path("..",
                       "..",
                       "static",
                       "onedegreewarmer"),
           showWarnings = FALSE)

temperatures_xlsx_location <-
  file.path("..", "..", "static",
            "onedegreewarmer",
            "temperatures.xlsx")

temperaturesMerged_csv_location  <-
  file.path("..", "..",
            "static",
            "onedegreewarmer",
            "temperaturesMerged.csv")

if (!file.exists(temperaturesMerged_csv_location)) {
  #gets each sheet of the excel file, and reads it separately
  temperaturesL <- lapply(excel_sheets(temperatures_xlsx_location),
                          read_excel,
                          path = temperatures_xlsx_location)
  names(temperaturesL) <- excel_sheets(temperatures_xlsx_location)
  temperatures <- purrr::map_df(.x = temperaturesL,
                                .f = bind_cols,
                                .id = "City") %>% 
    select(-2)

  write_csv(x = temperatures,
            path = temperaturesMerged_csv_location)
} else {
  temperatures <- read_csv(temperaturesMerged_csv_location)
}

```


```{r message=FALSE}


indicators <- read_csv(file = file.path("..",
                                        "..",
                                        "static",
                                        "onedegreewarmer",
                                        "indicators.csv"))

tempCoord <- left_join(x = temperatures,
                       y = indicators %>%
                         select(City = city, lat, lon, size),
                       by = "City")

 
```


```{r staticOneCity}
i <- "Granada"

oneCity <- 
  temperatures %>% 
  filter(City == i) %>% 
  mutate(century = str_extract(string = date,
                               pattern = "[[:digit:]][[:digit:]]")) %>% 
  group_by(century) %>% 
  mutate(AverageTempCentury = mean(temperature)) %>% 
  mutate(date = lubridate::year(date)) %>% 
  rename(Date = date)

oneCity <-  oneCity %>% 
  mutate(TempDifference = max(oneCity$AverageTempCentury)-min(oneCity$AverageTempCentury)) 


staticOneCity <- 
  ggplot(data = oneCity,
         mapping = aes(x = Date,
                       y = temperature,
                       colour = `temperature`,
                       AverageTempCentury)) +
  geom_line(size = 1, color = "gray") +
  geom_point() +

  # 21st century horizontal line
  geom_segment(aes(x = 1900,
                   y = unique(oneCity$AverageTempCentury)[1],
                   xend = 2000,
                   yend = unique(oneCity$AverageTempCentury)[1]),
               colour = "#6f2c91",
               linetype=2) +
  geom_segment(aes(x = 2000,
                   y = unique(oneCity$AverageTempCentury)[1],
                   xend = 2017,
                   yend = unique(oneCity$AverageTempCentury)[1]),
               colour = "#a6ce39",
               linetype=1) +
  # 20th century horizontal line
  geom_segment(aes(x = 1900,
                   y = unique(oneCity$AverageTempCentury)[2],
                   xend = 2000,
                   yend = unique(oneCity$AverageTempCentury)[2]),
               colour = "#a6ce39") +
  geom_segment(aes(x = 2000,
                   y = unique(oneCity$AverageTempCentury)[2],
                   xend = 2017,
                   yend = unique(oneCity$AverageTempCentury)[2]),
               colour = "#6f2c91",
               linetype=2) +
  # define colour of points
  scale_color_viridis_c(option = "magma",
                        direction = -1,
                        guide = FALSE) + 
  # define breaks and labeling on scales
  scale_x_continuous(name = "",
                     breaks = c(1900, 1925, 1950, 1975, 2000, 2017)) +
  scale_y_continuous(name = "",
                     labels = function(x) scales::number(x = x, suffix = "°")) +
  # basic styling and font
  theme_minimal(base_family = "Carlito",
                base_size = 16)+
  # labels for horizontal lines
  annotate(geom = "text",
           x = 1915,
           y = unique(oneCity$AverageTempCentury),
           label = c("19th century average",
                     "20th century average"),
           vjust = -1,
           fontface = 2,
           size = 4) +
  labs(title = paste("Average temperature in", i, "(1900-2017)"), 
       caption = "Source: EdjNet/ECMWF") +
  theme(plot.caption=element_text(hjust = 0,
                                  colour = "darkgray")) +
    coord_cartesian(clip = 'off') +
# label final temperatures 
  geom_label_repel(data = oneCity %>%
                     distinct(century, .keep_all = TRUE) %>%
                     select(-Date),
                   mapping = aes(x = 2000,
                                 y = AverageTempCentury,
                                 label = scales::number(AverageTempCentury,
                                                        accuracy = 0.01,
                                                        suffix = "°")),
                   nudge_x = c(3, -3),
                   nudge_y = c(-1,3),
                   segment.size  = 0.2,
                   colour = "black") +
  # draw curved line
  geom_curve(aes(x = 2000,
                 y = unique(oneCity$AverageTempCentury)[1],
                 xend = 2017,
                 yend = unique(oneCity$AverageTempCentury)[2]),
             colour = "#6f2c91",
             linetype=1,
             curvature = 0.6,
             arrow =  arrow(length = unit(0.15, "inches"))) +
   geom_label(data = oneCity %>%
                distinct(TempDifference, .keep_all = TRUE),
              mapping = aes(x = 2020,
                            y = mean(AverageTempCentury),
                            label = scales::number(TempDifference,
                                                   accuracy = 0.01,
                                                   suffix = "°")), colour = "black")
    
staticOneCity
```


## Animate the graph

Thanks to the excellent work done by [Thomas Lin Pedersen](https://twitter.com/thomasp85/) on `gganimate`, it's now a single line of code to add animation to the above graph. 

```{r}
animatedOneCity <- staticOneCity +
  transition_reveal(id = City, along = Date)
```

And here it is as a GIF: 

```{r animatedOneCity, message=FALSE}
animate(plot = animatedOneCity +
  transition_reveal(id = City, along = Date),
 width = 800, height = 600, res = 120)
```

# Export as high quality video

Thanks to the latest `av` packge by rOpenSci, it's now easy to export the graph in high quality video formats ([see the blog bost announcing the release of the package](https://ropensci.org/technotes/2018/10/06/av-release/)). 

```{r eval = FALSE}

pacman::p_load("av") # for video rendering

# Render and show the video
mp4 <- animate(animatedOneCity, renderer = av_renderer('animation.mp4'),
	width = 1920, height = 1080, res = 200, fps = 25)
utils::browseURL('animation.mp4')
```

# Bulk create graphs

[link to repository with script coming soon]
